Mergeable Types

### 2017-02-27 paper/model.tex
edits

Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
quadrant contains multiple pixels of different colors. Quadrants are
expanded into a tree structures as and when pixels are colored.  The
representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
and \C{deriving versioned} directive are explained below.

\begin{figure}
\centering
\includegraphics[scale=0.5]{Figures/canvas-merging}

\caption{Merging concurrent versions (\C{v1} and \C{v2} ) of a drawing
canvas. \C{lca} is their common ancestor.}
\label{fig:canvas-merging}
\end{figure}

\drawsome lets multiple users collaborate on a canvas that is
conceptually shared among them. Under a shared-memory abstraction,
there would be a single copy of the canvas that is updated
concurrently by multiple clients; from the perspective of any single
client, the canvas could change without any explicit
intervention. \name ascribes functional semantics to sharing by
letting each client work on its own version of the state (the tree
data structure), later merging concurrent versions on-demand.  Thus,
the primary artifact of the \name programming model is a versioned
data structure in which different versions are managed by different
clients.

@@

The \drawsome example demonstrates the utility of mergeable data types
and \name programming model in building concurrent applications with
conceptual sharing of state. The model lets concurrent computations be
composed around any mergeable type. As exemplified by \drawsome,
writing a three-way \C{merge} function is the only creative process in
lifting an OCaml data type to a mergeable type. We now present few
more examples of mergeable types, and isolate a recurring pattern in
their merge functions that serves as a guide to write merge functions
for more sophisticated types.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}

  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{figure}

@@

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
\C{i}'th position with \C{x}. Lists of mergeable items are mergeable
while preserving the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ also in the merged list.
  \item If an element $y$ substitutes $x$ in one list, then $y$ also
  substitutes $x$ in the merged list (i.e., $x$ is absent and $y$ is
  present), unless $x$ is deleted or substituted in the other list. In
  the former case, deletion wins, and neither $x$ nor $y$ occur in the
  merged list. In the latter case, if $x$ is substituted by a
  different element $z$ in the other list, then the merged list
  substitutes $x$ by a merge of $y$ and $z$, defined as per the merge
  semantics of the list element type.
\end{itemize}
Thus list is an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed. The compositionality is
reified in the the signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
  computes the shortest sequence of list operations that need to be
  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
  \emph{edit sequence}. The length of the sequence corresponds to the
  standard notion of the \emph{edit distance} between the two lists,
  which can be computed in polynomial time~\cite{wagner-fischer}.
  \C{edit\_seq} modifies one such algorithm to also compute the edit
  sequence (an \C{edit list}). The implementation specified in
  Fig.~\ref{fig:mlist-sig} represents edits using the type \C{edit}.
  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
  \C{delete}, and \C{subst}, respectively\footnote{
    Note that \C{I}, \C{D} and \C{S} are \emph{not} effects; no list
    function generates them, and they are not exchanged between
    concurrent threads. They are simply tags used by the list merge
    operation for convenience.
  }. The subst constructor also carries the \C{A.t} element that was
  substituted. The element serves as the \C{lca} argument to
  \C{A.merge} in case of concurrent conflicting substitutions.
  Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.
  The sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to
  \C{[c;a;b;s]} because \C{subst 3 s (insert c 0 [a;b;c]) =
  [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
  another edit sequence $s_2$. A notable transformation rule is for a
  substitution w.r.t another substitution at the same position. Since
  the substituting items could be different, the function relies on
  the item merge function (\C{A.merge}) to merge them into a single
  item. For instance, in Fig.~\ref{fig:list-eg}, there are
  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
  merged list therefore substitutes \C{c} with the merge of \C{d} and
  \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

\GK{Somewhere here, I want to say:} To better appreciate the virtues
of mergeable types, we encourage the reader to think about how one
might implement the mergeable list type with the explicit effect
representation described in Sec.~\ref{sec:intro}.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{Figures/shopping-list}

  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{figure}

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of merge semantics of the individual types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
An alternative instantiation of \C{MList} is for a quantity type
(\C{Qty}) that, like \C{Int}, supports \C{add} and \C{subtract}
operations, but defines its \C{merge} semantics a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.  is given
below.

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application
allows its users to collaboratively build a grocery shopping
list (listing items in decreasing ), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}). Merge
function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) with same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items
(i.e., \C{MList(Item)}). \C{MList(Item).merge} automatically lifts the
merge semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates. Here, two users, Alice and
Bob, concurrently edit a shopping list whose initial contents are
as milk and eggs. While both simultaneously update the quantity of
eggs, Bob also removes milk and inserts candy. The merged shopping
list contains eggs and candy, where the quantity of eggs is obtained
by merging Alice's and Bob's quantity as per \C{Qty.merge} semantics.

>>>>

Fig.~\ref{fig:canvas-sig} shows the signature of the \drawsome
application. \drawsome represents a free-hand drawing canvas in terms
of a tree of quadrants.  Each quadrant is simply a leaf node
containing a single pixel (\C{r},\C{g}, or \C{b}) or a tree, if the
quadrant contains multiple pixels of different colors. Quadrants are
expanded into a tree structures as and when pixels are colored.  The
representation is thus optimized for sparse canvases, such as
whiteboards. The application supports three simple operations:
creating a new canvas, setting the pixel at a specified coordinate,
and returning the pixel at a given coordinate. The \C{merge} function
and \C{deriving versioned} directive are explained below.  \drawsome
lets multiple users collaborate on a canvas that is conceptually
shared among them. Under a shared-memory abstraction, there would be a
single copy of the canvas that is updated concurrently by multiple
clients; from the perspective of any single client, the canvas could
change without any explicit intervention. \name ascribes functional
semantics to sharing by letting each client work on its own version of
the state (the tree data structure in this example), later merging
concurrent versions on-demand.  Thus, the primary artifact of the
\name programming model is a versioned data structure in which
different versions are managed by different clients.

@@

The \drawsome example demonstrates the utility of mergeable data types
and the \name programming model in building concurrent applications
with conceptual state sharing. The model lets concurrent computations
be composed around any mergeable type. As exemplified by \drawsome,
writing a three-way \C{merge} function is the only creative process
involved to lift an OCaml data type to a mergeable type. We now
present some more examples of mergeable types, and isolate a recurring
pattern in their merge definitions that serves as a guide to write
merge functions for more sophisticated types.

@@

{\bf Lists}.  As an abstract data type, lists support three
operations: \C{insert x i l}, to insert an element \C{x} at position
\C{i} in the list \C{l}, \C{delete i l}, to delete the element at
\C{i}'th position, and \C{subst i x l}, to substitute the element at
the \C{i}'th position with \C{x}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike Amazon's replicated shopping cart~\cite{dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}
Thus, lists serve as an example of a polymorphic mergeable data type that
demonstrates how mergeability can be composed.  Compositionality is
reified in the the signature of a mergeable list library as shown in
Fig.~\ref{fig:mlist-sig}; observe that the type of items in the list
(\C{A.t}) is also required to be mergeable (\C{A.merge}).

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml included in the appendix.
The functions are described below.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
    computes the shortest sequence of list operations that need to be
    applied on \C{v} to obtain \C{v'}. Such a sequence is called an
    \emph{edit sequence}. The length of the sequence corresponds to
    the standard notion of the \emph{edit distance} between the two
    lists, which can be computed in polynomial
    time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such
    algorithm to also compute the edit sequence (an \C{edit
      list}). The implementation specified in Fig.~\ref{fig:mlist-sig}
    represents edits using the type \C{edit}.  Constructors \C{I},
    \C{D}, and \C{S} stand for \C{insert}, \C{delete}, and \C{subst},
    respectively.\footnote{ Note that \C{I}, \C{D} and \C{S} are
      \emph{not} effects; no list function generates them, and they
      are not exchanged between concurrent threads. They are simply
      tags used by the list merge operation for convenience.  Indeed,
      to better appreciate the virtues of mergeable types, we
      encourage the reader to think about how one might implement a
      replicated list with this functionality using an explicit effect
      representation as described in Sec.~\ref{sec:intro}.}  The subst
    constructor also carries the \C{A.t} element that was
    substituted. The element serves as the \C{lca} argument to
    \C{A.merge} in case of concurrent conflicting substitutions.
    Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample
    list.  The sequence \C{[I(c,0); S(3,c,s)]} maps the list
    \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
      [a;b;c]) = [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(1,c,d)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} (Fig.~\ref{fig:mlist-sig}) computes such
  \emph{operational transformations} for an edit sequence $s_1$ w.r.t
  another edit sequence $s_2$. A notable transformation rule is for a
  substitution w.r.t another substitution at the same position. Since
  the substituting items could be different, the function relies on
  the item merge function (\C{A.merge}) to merge them into a single
  item. For instance, in Fig.~\ref{fig:list-eg}, there are
  simultaneous substitutions to the item \C{c} in $s_1$ and $s_2$;
  while $s_1$ substitutes it with \C{d}, $s_2$ does it with \C{s}. The
  merged list therefore substitutes \C{c} with the merge of \C{d} and
  \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

A significant advantage of mergeable types is their composability. As
demonstrated by \C{MList}, the merge semantics of a composed mergeable
type is a composition of the merge operations of its constituent types.  For
instance, a mergeable list of integers can be obtained by
instantiating \C{MList} for an \C{Int} type, where \C{Int.merge} is
the same as \C{Counter.merge} (Sec.~\ref{sec:intro}). 
An alternative instantiation of \C{MList} is for a quantity type
\begin{wrapfigure}{l}{.45\textwidth}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}
  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{wrapfigure}
(\C{Qty}) that supports \C{add} and \C{subtract}
operations, but defines its \C{merge} opeation a little differently:
\begin{ocaml}
let merge lca v1 v2 = if v1 = lca then v2 else (if v2 = lca then v1 else max v1 v2)
\end{ocaml}
\C{MList(Qty)}, like \C{MList(Int)}, is a list of integers, and has
the same semantics under a sequential execution. Their differences
however manifest under a concurrent execution, where both adopt
different methods to reconcile concurrent updates.

@@

{\bf Shopping List}. A non-trivial demonstration of composability of
mergeable types is a shopping list application. The application allows
its users to collaboratively build a grocery shopping list (listing
items by decreasing quantity), by adding new items with a specified
quantity, deleting existing items, and increasing or decreasing their
quantity. An item (\C{Item.t}) is represented as a tuple of its name
(a string) and the quantity (A \C{Qty.t}). The merge function for
\begin{wrapfigure}{l}{.5\textwidth}
  \centering
  \includegraphics[scale=0.3]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
\end{wrapfigure}
\C{Item.t} merges the quantities of items (using \C{Qty.merge}) having
the same names. Items with different names are considered distinct. A
shopping list is a mergeable list of items (i.e.,
\C{MList(Item)}). \C{MList(Item).merge} automatically lifts the merge
semantics for items to merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update the
quantity of eggs, Bob also removes milk and inserts candy. The merged
shopping list contains eggs and candy, where the quantity of eggs is
obtained by merging Alice's and Bob's quantity as per the definition
of \C{Qty.merge}.
